Economics and Crime Patterns: The Interplay Between CPI Changes and Vehicle Theft Frequencies

### 2024-04-19 paper/paper.qmd
fixed typo in final version

In recent years, rapid global economic development has accelerated currency devaluation. During the pandemic, constrained economic output had led to significant job losses and a shortage in global supply chains. These factors have contributed to a sharp rise in inflation rates across Canada. Data from Statistics Canada 2024 [@statcanCPI] indicates that these changes have made it increasingly difficult for people to afford their daily expenses. Meanwhile, regional crime rates have also significantly increased. According to the Canada Newswire [@equite2023], auto theft trends in Ontario rose dramatically by 48.2%, with a vehicle now stolen every five minutes between 2021 and 2023.

@@

At the same time, the economic strain from the pandemic led to a significant increase in vehicle thefts, mirroring the rise in the CPI. Vehicle thefts spiked notably in Toronto, reflecting trends observed in several urban areas during the pandemic, as discussed in the CBC News article [@cbc2020cartheft]. This surge was particularly evident in categories essential for daily life, suggesting that financial difficulties may push some individuals toward crime. Similar trends have been observed globally, a study by [@northeastern2020crime] details how the pandemic has influenced crime rates across the United States, noting a significant increase in violent crimes such as homicides and assaults during 2020. Lockdowns and changes in public behaviour have altered typical crime patterns, affecting when and where crimes occurred.

@@

While discussing the pandemic's impact, this study also examines the relationship between various CPI categories and vehicle theft rates in Toronto over the past ten years. Similar to findings from research on the economic impacts of the fracking boom [@cgoFracking2020], our analysis indicates complex interactions between economic conditions and crime rates. We found that increases in the 'All_items' CPI correspond with higher vehicle theft rates. This suggests that as the cost of basic goods and services increases, so does the frequency of vehicle thefts, potentially as individuals face greater financial stress. Conversely, we observe that higher spending in non-essential areas like recreation and education is associated with fewer thefts, suggesting that such discretionary expenditures might help reduce crime.

>>>>

In recent years, rapid global economic development has accelerated currency devaluation. During the pandemic, constrained economic output had led to significant job losses and a shortage in global supply chains. These factors have contributed to a sharp rise in inflation rates across Canada. Data from Statistics Canada [@statcanCPI] indicates that these changes have made it increasingly difficult for people to afford their daily expenses. Meanwhile, regional crime rates have also significantly increased. According to the Canada Newswire [@equite2023], auto theft trends in Ontario rose dramatically by 48.2%, with a vehicle now stolen every five minutes between 2021 and 2023.

@@

At the same time, the economic strain from the pandemic led to a significant increase in vehicle thefts, mirroring the rise in the CPI. Vehicle thefts spiked notably in Toronto, reflecting trends observed in several urban areas during the pandemic, as discussed in the CBC News article [@cbc2020cartheft]. This surge was particularly evident in categories essential for daily life, suggesting that financial difficulties may push some individuals toward crime. Similar trends have been observed globally, a study by Northeastern University [@northeastern2020crime] details how the pandemic has influenced crime rates across the United States, noting a significant increase in violent crimes such as homicides and assaults during 2020. Lockdowns and changes in public behaviour have altered typical crime patterns, affecting when and where crimes occurred.

@@

While discussing the pandemic's impact, this study also examines the relationship between various CPI categories and vehicle theft rates in Toronto over the past ten years. Similar to findings from research on the economic impacts of the fracking boom [@cgo2020fracking], our analysis indicates complex interactions between economic conditions and crime rates. We found that increases in the 'All_items' CPI correspond with higher vehicle theft rates. This shows that as the cost of basic goods and services increases, so does the frequency of vehicle thefts, potentially as individuals face greater financial stress. Conversely, we observe that higher spending in non-essential areas like recreation and education is associated with fewer thefts, suggesting that such discretionary expenditures might help reduce crime.
